% !TEX program = lualatex

\documentclass[letterpaper]{article}
\title{Pontifical Mass at the Faldstool\\\small Without the Assisting Priest}

\usepackage{rubrics}


\usepackage{csquotes}
\usepackage[backend=biber,style=reading,style=verbose-ibid]{biblatex}
\addbibresource{~/Library/texmf/bibtex/bib/rubrics.bib}
%\bibliography{rubrics}

\begin{document}
\maketitle


\section{Vesting in the Sactisty}


\section{Confession, Introit, Gloria}

\begin{enumerate} 

    \item D and SD to the left of B, MC1 to his right\footnote{MC1 is taking
        the place of AP.}, MC2 to the right of SD.

    \item All bow to the choir, D removes miter, SD hands book of Gospels to
        MC2, and Mass begins.

    \item SD imposes maniple at \textit{Indulgéntiam}.

    \item After kissing the altar, B turns to his left and kisses the book of
        Gospels, held by SD, the text pointed out by MC1. MC2 takes the book
        from SD and goes to the Epistle side to prepare book- and candle-bearers.

    \item B incenses the altar assisted by D and SD. SD\footnote{S.R.C. n.
        2138} imposes miter and D incenses B with three swings.

    \item B receives crozier, turns to bow to the cross, and then goes
        \textit{per breviorum} to the faldstool, flanked by D and SD.

    \item B sits for a few moments, then crozier away, D removes miter, B turns
        to the left to face the altar and reads the introit. Candle-bearer
        stands to the right of the book bearer. D stands to the right of B, SD
        to his left\footcite[See note 5, p. 241,][]{stehle}.

    \item B recites \textit{Kyrie} with the ministers, turns to the right to
        face the people, sits and receives miter and gremiale from D. D and SD bow to B and sit.

    \item Minsters rise after the last \textit{Kyrie}\footcite[][227, p.
        241]{stehle}, gremiale and miter removed by D; D and SD stand
        \textit{unum post alium}. B intones the \textit{Gloria}; D and SD flank
        B for the recitation. B and ministers sit for the remainder of the sung
        \textit{Glória}.

    \item At the end of the \textit{Glória} B and ministers rise, D and SD
        stand \textit{unum post alium} and B sings \textit{Dóminum vobíscum.} B
        sings the orations. SD receives the Book of Epistles from MC2 after the
        last \textit{Per Dóminum}, and stands with MC2 some distance from B.

\end{enumerate}

\section{Epistle and Gospel}

\begin{enumerate}

    \item B and D sit. SD bows to B and sings the Epistle; afterwards he
        repeats the reverence, goes the middle and genuflects, and kneels to
        receive the bishop's blessing.

    \item 

\end{enumerate}

%\bibliographystyle{plain}
\printbibliography
    
\end{document}
