% !TEX program = LuaLaTeX
\documentclass[11pt]{amsbook}

\usepackage{geometry}
%showframe
\geometry{
	paperheight=8.5in,
	paperwidth=5.5in,
	top=0.5in,
	bottom=0.5in,
	inner=0.5in,
	outer=0.5in,
}
\usepackage{layout}
\usepackage[letterspace=500]{microtype}
\usepackage[T1]{fontenc}
\usepackage[usefilenames,RMstyle={Text,Semibold},SSstyle={Text,Semibold},TTstyle={Text,Semibold},DefaultFeatures={Ligatures=Common}]{plex-otf} %
\usepackage[latin]{babel}
\usepackage{fontspec}
\usepackage{ragged2e}
\usepackage{setspace}
\usepackage{xcolor}
\usepackage{stackengine}
\usepackage{tabularx}
\usepackage{longtable}
\renewcommand\stacktype{L}
\renewcommand\stackalignment{c}
\setstackgap{L}{8pt}

\newcommand{\subword}[2]{%
	\noindent
	\begin{justifying}
	\stackunder{\large\ #1}{\tiny\ #2}%
	\end{justifying}
	}

\newcommand{\priest}[1]{%
	\begin{raggedright}
		\textit{\noindent\footnotesize #1}
	\end{raggedright}
	}

\newcommand{\centerRubric}[1]{%
	\begin{center}
	\noindent
	\footnotesize
	\color{red}
	\textit{S. }
	#1
	\end{center}
	}

\newcommand{\centerRubricAlt}[1]{%
	\begin{center}
	\noindent
	\footnotesize
	\color{red}
	#1
	\end{center}
	}


\newcommand{\rubric}[1]{%
	\begin{footnotesize}
	\color{red}
	#1
	\end{footnotesize}
	}


\newcommand{\server}[1]{%
	\noindent
	#1
	}

\newcommand{\p}{%
	\noindent
	\textit{\color{red}\small P.}
	}
	
\newcommand{\s}{%
	\noindent
	\textit{\color{red}\small S.}
	}

\newcommand{\DV}{%
	\noindent
	\p \priest{Dóminus vobíscum.} \s \server{Et cum spíritu tuo.}
	}

\newcommand{\pp}{%
	\textit{\color{red}\small P.*}
	}

\newcommand{\sss}{%
	\textit{\color{red}\small S.*}
	}

\pagestyle{plain}
%\pagenumbering{gobble}

\begin{document}
    \stackText
    \setlength\tabcolsep{0pt}
    %\layout
    
    \begin{center}
        \Huge\MakeUppercase{Mass Server's Card}
        \bigskip\hrule\smallskip
        \textls{\MakeUppercase{\small Most Holy Trinity Seminary}}
        \smallskip\hrule
    \end{center}
    
    \begin{longtable}{@{}l@{}p{4.32in}@{}}
        \multicolumn{2}{c}{\footnotesize\color{red}\textit{S.} Blesses himself with the priest.} \\
        \p   & \priest{In nómine Patris, {\color{red} \maltese} et Fílii, et Spíritus Sancti. Amen.} \\
        \pp  & \priest{Introíbo ad altáre Dei.} \\
        \sss & \server{\subword{Ad}{Ahd} \subword{Deum}{Day´oom} \subword{qui}{kwee} \subword{l\ae tíficat}{lay-tee´-fee-kaht} \subword{juventútem}{yoo-ven-too´-tem} \subword{meam.}{may´ahm.}} \\
        \pp  & \priest{Júdica me, Deus, ... et dolóso érue me.} \\
        \sss & \server{\subword{Quia}{Quee´ah} \subword{tu}{too} \subword{es,}{es,} \subword{Deus,}{Day´oos,} \subword{fortitúdo}{for-tee-too´doh} \subword{mea:}{may´ah:} \subword{quare}{quah´ray} \subword{me}{may} \subword{repulísti,}{ray-poo-lee´stee,} \subword{et}{ett} \subword{quare}{quah´ray} \subword{tristis}{tree´stees} \subword{incédo,}{in-chay´doh,} \subword{dum}{doom} \subword{afflígit}{ahf-flee´jeet} \subword{me}{may} \subword{inimícus?}{in-ee-mee´koos?}} \\
        \pp  & \priest{Emítte lucem tuam ... in tabernácula tua.} \\
        \sss & \server{\subword{Et}{Et} \subword{introíbo}{in-troh-ee´boh} \subword{ad}{ahd} \subword{altáre}{ahl-tah´ray} \subword{Dei:}{Day´ee:} \subword{ad}{ahd} \subword{Deum}{Day´oom} \subword{qui}{kwee} \subword{l\ae tíficat}{lay-tee´-fee-kaht} \subword{juventútem}{yoo-ven-too´-tem} \subword{meam.}{may´ahm.}} \\
        \pp  & \priest{Confitébor tibi in cíthara, ... quare contúrbas me?} \\
        \sss & \server{\subword{Spera}{Spay´rah} \subword{in}{in} \subword{Deo,}{Day´oh} \subword{quóniam}{quoh´nee-ahm} \subword{adhuc}{ahd´hook} \subword{confitébor}{kon-fee-tay´bore} \subword{illi:}{eel´lee:} \subword{salutáre}{sah-loo-tah´ray} \subword{vultus}{vool´toos} \subword{mei,}{may´ee,} \subword{et}{ett} \subword{Deus}{Day´oos} \subword{meus.}{may´oos.}} \\
        \multicolumn{2}{c}{\footnotesize\color{red}\textit{S.} Bows his head with the priest.} \\
        \pp  & \priest{Glória Patri, et Fílio, et Spirítui Sancto.} \\
        \sss & \server{\subword{Sicut}{See´koot} \subword{erat}{air´rot} \subword{in}{in} \subword{princípio,}{prin-chee´pee-oh} \subword{et}{ett} \subword{nunc,}{noonk,} \subword{et}{ett} \subword{semper:}{sem´pair:} \subword{et}{ett} \subword{in}{in} \subword{s\'\ae cula}{say´koo-lah} \subword{sæculórum.}{say-koo-lore´room.} \subword{Amen.}{Ah´men.}} \\
        \multicolumn{2}{l}{\tiny\color{red} *Omitted at Masses for the Dead.} \\
    	\vspace{-0.46in}    
    \end{longtable}
    
    \begin{longtable}{@{}l@{}p{4.37in}@{}}
        \p & \priest{Introíbo ad altáre Dei.} \\
        \s & \server{\subword{Ad}{Ahd} \subword{Deum}{Day´oom} \subword{qui}{kwee} \subword{l\ae tíficat}{lay-tee´-fee-kaht} \subword{juventútem}{yoo-ven-too´-tem} \subword{meam.}{may´ahm.}} \\
        \multicolumn{2}{c}{\footnotesize\color{red}\textit{S.} Blesses himself with the priest.} \\
        \p & \priest{Adjutórium nostrum {\color{red} \maltese} in nómine Dómini.} \\
        \s & \server{\subword{Qui}{Quee} \subword{fecit}{fay´cheet} \subword{cælum}{chay´loom} \subword{et}{ett} \subword{terram.}{tair´rahm.}} \\
        \p & \priest{Confíteor ... ad Dóminum Deum nostrum.} \\
    \end{longtable}
    
    \centerRubric{Bows towards the priest and says:}
    
    \begin{longtable}{@{}l@{}p{4.37in}@{}}
        \s & \server{\subword{Misereátur}{Miz-zay-ray-ah´toor} \subword{tui}{too´ee} \subword{omnípotens}{ohm-nee´poh-tenz} \subword{Deus,}{Day´oos,} \subword{et}{ett} \subword{dimíssis}{dee-mee´sees} \subword{peccátis}{pay-kah´tees} \subword{tuis,}{too´ees,} \subword{perdúcat}{pair-doo´kott} \subword{te}{tay} \subword{ad}{ahd} \subword{vitam}{vee´tahm} \subword{ætérnam.}{ay-tair´nahm.}} \\
        \p & \priest{Amen.} \\    
    \end{longtable}
    
    \centerRubric{Bows with head and shoulders towards the altar and says:}
    
    \begin{longtable}{@{}l@{}p{4.37in}@{}}    
        \s & \server{\subword{Confíteor}{Kon-fee´tay-ore} \subword{Deo}{Day´oh} \subword{omnipoténti,}{ohm-nee-poh-ten´tee,} \subword{beát\ae}{bay-ah´toh} \subword{Marí\ae}{Mah-ree´ay} \subword{semper}{sem´pair} \subword{Vírgini,}{Veer´jee-nee,} \subword{beáto}{bay-ah´toh} \subword{Michaéli}{Mee-kah-ay´lee} \subword{Archángelo,}{Ark-ahn´jay-loh,} \subword{beáto}{bay-ah´toh} \subword{Joánni}{Yoh-ahn´nee} \subword{Baptíst\ae,}{Bahp-tee´stay,} \subword{sanctis}{sahnk´tees} \subword{Apóstolis}{Ah-poh´stoh-lis} \subword{Petro}{Pay´troh} \subword{et}{ett} \subword{Paulo,}{Pow´loh,} \subword{ómnibus}{ohm´nee-boos} \subword{Sanctis,}{Sahnk´tees,} \subword{et}{ett} \subword{tibi,}{tee´bee,} \subword{pater:}{pah´tair:} \subword{quia}{que´ah} \subword{peccávi}{pay-kah´vee} \subword{nimis}{nee´mees} \subword{cogitatióne,}{koh-jee-tahd-see-oh´nay} \subword{verbo}{vair´boh} \subword{et}{ett} \subword{ópere:}{oh´pay-ray:} \subword{mea}{may´ah} \subword{culpa,}{kool´pah,} \subword{mea}{may´ah} \subword{culpa,}{kool´pah,} \subword{mea}{may´ah} \subword{máxima}{mahk´see-mah} \subword{culpa.}{keel´pah.} \subword{Ideo}{Ee´day-oh} \subword{precor}{pray´kore} \subword{beátam}{bay-ah´tam} \subword{Maríam}{Mah-ree´ahm} \subword{semper}{sem´pair} \subword{Vírginem,}{Veerj́ee-nem,} \subword{beátum}{bay-ah´tuum} \subword{Michaélem}{Mee-kah-ay´lem} \subword{Archángelum,}{Ark-ahn´jay-loom,} \subword{beátum}{bay-ah´toom} \subword{Joánnem}{Yo-ahn´nem} \subword{Baptístam,}{Bahp-tee´stahm,} \subword{sanctos}{sahnk´tohs} \subword{Apóstolos}{Ah-poh´stoh-lohs} \subword{Petrum}{Pay´troom} \subword{et}{ett} \subword{Paulum,}{Pow´loom,} \subword{omnes}{ohm´nays} \subword{Sanctos,}{Sahnk´tohs} \subword{et}{ett} \subword{te,}{tay,} \subword{pater,}{pah´tair,} \subword{oráre}{oh-rah´ray} \subword{pro}{proh} \subword{me}{may} \subword{ad}{ahd} \subword{Dóminum,}{Doh´mee-noom,} \subword{Deum}{Day´oom} \subword{nostrum.}{noh´stroom.}} \\
        \p & \priest{Misereátur ... \ae térnam.} \s \server{Amen.} \\
        \p & \priest{Indulgéntiam ... Dóminus.} \s \server{Amen.} \\
        \multicolumn{2}{c}{\footnotesize\color{red}For a bishop, \textit{S.} presents the maniple.} \\
        \p & \priest{Deus, tu convérsus vivificábis nos.} \\
        \s & \server{\subword{Et}{Ett} \subword{plebs}{playbs} \subword{tua}{too´ah} \subword{lætábitur}{lat-tah´bee-toor} \subword{in}{in} \subword{te.}{tay.}} \\
        \p & \priest{Osténde nobis, Dómine, misericórdiam tuam.} \\
        \s & \server{\subword{Et}{Ett} \subword{salutáre}{sah-loo-tah´ray} \subword{tuum}{too´oom} \subword{da}{dah} \subword{nobis.}{noh´bees.}} \\
        \p & \priest{Dómine, exáudi oratiónem meam.} \\
        \s & \server{\subword{Et}{Ett} \subword{clamor}{clah´mohr} \subword{meus}{may´oos} \subword{ad}{ahd} \subword{te}{tay} \subword{véniat.}{vay´nee-aht.}} \\
        \p & \priest{Dóminus vobíscum.} \\
        \s & \server{\subword{Et}{Ett} \subword{cum}{koom} \subword{spíritu}{spee´ree-too} \subword{tuo.}{too´oh.}} \\
    \end{longtable}
    
    \centerRubricAlt{At the middle of the altar, after the reading of the \textit{Introit:}}
    
    \p \priest{K\'yrie, eléison.} \s \server{\subword{K\'yrie,}{Kee´ree-ay,} \subword{eléison.}{ay-lay´ee-sohn.}} 
    
    \p \priest{K\'yrie, eléison.} \s \server{\subword{Christe,}{Kree´stay} \subword{eléison.}{ay-lay´ee-sohn.}} 
    
    \p \priest{Christe, eléison.} \s \server{Christe, eléison.} \p \priest{K\'yrie, eléison.} 
    
    \s \server{K\'yrie, eléison.} \p \priest{K\'yrie, eléison.}
    
    \centerRubricAlt{After the \textit{K\'yrie eléison} or after the \textit{Glória} if this is said:}
    
    \DV
    
    \pagebreak
    
    \centerRubricAlt{Or, for a bishop, after the \textit{Glória} is said:}
    
    \p \priest{Pax vobis.} \s \server{Et cum spíritu tuo.}
    
    \p \priest{Orémus, etc.} \s \server{Amen.}
    
    \p \priest{Orémus, etc.}
    
    \s \server{Amen.} -- \rubric{After the last prayer when there are more than one.}
    
    \rubric{After the Epistle: }\s \server{\subword{Deo}{Day´oh} \subword{grátias.}{grahd´see-ahs.}}
    
    \centerRubricAlt{At the beginning of the Gospel:}
    
    \DV
    
    \p \priest{Sequéntia Sancti Evangélii, etc.}
    
    \s \server{\subword{Glória}{Gloh´ree-ah} \subword{tibi,}{tee´bee,} \subword{Dómine.}{Doh´mee-nay.}}
    
    \rubric{After the Gospel: }\s \server{\subword{Laus}{Louse} \subword{tibi,}{tee´bee,} \subword{Christe.}{Kree´stay.}}
    
    \centerRubricAlt{After the Gospel or after the \textit{Credo} if this is said:}
    
    \DV
    
    \centerRubricAlt{After the \textit{Lavábo} the priest says towards the people:}
    
    \begin{longtable}{@{}l@{}p{4.37in}@{}}
    
    \p & \priest{Oráte, fratres ... Patrem onmipoténtem.} \\
    \s & \server{\subword{Suscípiat}{Soo-shee´pee-ott} \subword{Dóminus}{Doh´mee-noos} \subword{sacrifícium}{sah-kree-fee´chee-oom} \subword{de}{day} \subword{mánibus}{mah´nee-boose} \subword{tuis,}{too-eese,} \subword{ad}{ahd} \subword{laudem}{lou´dem} \subword{et}{ett} \subword{glóriam}{gloh´ree-ahm} \subword{nominis}{noh´mee-neese} \subword{sui,}{soo´ee,} \subword{ad}{ahd} \subword{utilitátem}{oo-tee-lee-tah´tem} \subword{quoque}{quoh´quay} \subword{nostram,}{noh´strahm,} \subword{totiúsque}{tohd-see-oos´quay} \subword{Ecclésiæ}{at-clay´see-ay} \subword{suæ}{soo´ay} \subword{sanctæ.}{sahnk´tay.}} \\
    
    \end{longtable}
    
    \centerRubricAlt{At the Preface}
    
    \p \priest{Per ómnia sǽcula sæculórum.} \s \server{Amen.}
    
    \DV
    
    \p \priest{Sursum corda.}
    
    \s \server{\subword{Habémus}{hah-bay´moos} \subword{ad}{ahd} \subword{Dóminum.}{Doh´mee-noom.}}
    
    \p \priest{Grátias agámus Dómino, Deo nostro.}
    
    \s \server{\subword{Dignum}{Deen´yoom} \subword{et}{ett} \subword{justum}{yoo´stoom} \subword{est.}{est.}}
    
    \centerRubricAlt{Before the \textit{Pater Noster} the priest genuflects and says:}
    
    \p \priest{Per ómnia sǽcula sæculórum.} \s \server{Amen.}
    
    \centerRubricAlt{At the end of the \textit{Pater Noster}:}
    
    \s \server{\subword{Sed}{Sed} \subword{líbera}{lee´bay-rah} \subword{nos}{nohs} \subword{a}{ah} \subword{malo.}{mah´loh.}}
    
    
    
    \centerRubricAlt{After the breaking of the Sacred Host:}
    
    \begin{longtable}{@{}l@{}p{4.37in}@{}}
    
    \p \priest{Per ómnia sǽcula sæculórum.}     & \s \server{Amen.}               \\
    
    \p \priest{Pax Dómini sit semper vobíscum.} & \s \server{Et cum spíritu tuo.} \\
    
    \end{longtable}
    
    \centerRubricAlt{As the priest takes up the chalice, if Communion is distributed:}
    
    \s \server{Confíteor, etc.} \rubric{(See page 2)}
    
    %\begin{longtable}{@{}l@{}p{4.37in}@{}}
    
    \p \priest{Misereátur ... \ae térnam.} \s \server{Amen.}
    
    \p \priest{Indulgéntiam ... Dóminus.}  \s \server{Amen.}
    
    %\end{longtable}
    
    \centerRubricAlt{At the center of the altar after the ablutions.}
    
    \DV
    
    \p \priest{Orémus, etc.} \s \server{Amen.}
    
    \p \priest{Orémus, etc.}
    
    \s \server{Amen.} -- \rubric{After the last prayer when there are more than one.}
    
    \centerRubricAlt{Returning to the center of the altar:}
    
    \begin{longtable}{@{}l@{}p{4.37in}@{}}
        \p \priest{Dóminus vobíscum.}                   & \s \server{Et cum spíritu tuo} \\
        \p \priest{Ite Missa est.}                      & \s \server{Deo grátias.}       \\
        \rubric{...or, if there is no \textit{Glória}:} &                                \\
        \p \priest{Benedicámus Dómino.}                 & \s \server{Deo grátias.}       \\
        \rubric{...or, in Masses for the Dead:}         &                                \\
        \p \priest{Requiéscant in pace.}                & \s \server{Amen.}              \\
    \end{longtable}
    
    \centerRubricAlt{Before the Blessing, for a Bishop:}
    
    \p \priest{Sit nomen Dómini benedíctum.}
    
    \s \server{\subword{Ex}{Eks} \subword{hoc}{ohk} \subword{nunc}{noonk} \subword{et}{ett} \subword{usque}{oos´quay} \subword{in}{in} \subword{s\'\ae culum.}{say´coo-loom.}}
    
    \p \priest{Adjutórium nostrum in nómine Dómini.}
    
    \s \server{Qui fecit c\ae lum et terram.}
    
    \centerRubricAlt{After the Blessing:}
    
    \p \priest{...Pater, et Fílius, et Spíritus Sanctus.} \s \server{Amen.}
    
    \centerRubricAlt{At the beginning of the last Gospel:}
    
    \begin{longtable}{@{}l@{}p{4.37in}@{}}
    \p \priest{Dóminus vobíscum.}          & \s \server{Et cum spíritu tuo}   \\
    \p \priest{Inítium or Sequéntia, etc.} & \s \server{Glória tibi, Dómine.} \\
    \end{longtable}
    
    \rubric{After the last Gospel:} \s \server{Deo grátias.}
    
\end{document}
